Visualisation of Brain Statistics with R-packages _ggseg_ and _ggseg3d_

### 2020-04-06 msc_ggseg.Rmd
update post-acceptance

---
title: Visualisation of Brain Statistics with R-packages _ggseg_ and _ggseg3d_
preprint: true
author: 
  - name: Athanasia M. Mowinckel
    affiliation: 1
    corresponding: true
    email: a.m.mowinckel@psykologi.uio.no
  - name: Didac Vidal-Piñeiro
    affiliation: 1
affiliation:
  - code: 1
    address: Center for Lifespan Changes in Brain and Cognition, University of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway
abstract: >
  There is an increased emphasis on visualizing neuroimaging results in more intuitive ways. Common statistical tools for dissemination, such as bar charts, lack the spatial dimension that is inherent in neuroimaging data. Here we present two packages for the statistical software R, _ggseg_ and _ggseg3d_, that integrate this spatial component. The _ggseg_ and _ggseg3d_ packages visualize pre-defined brain segmentations as both 2D polygons and 3D meshes, respectively. Both packages are integrated with other well-established R-packages, allowing great flexibility. In this tutorial, we present the main data and functions in the _ggseg_ and _ggseg3d_ packages for brain atlas visualization. The main highlighted functions are able to display brain segmentation plots in R. Further, the accompanying _ggsegExtra_-package includes a wider collection of atlases, and is intended for community-based efforts to develop more compatible atlases to _ggseg_ and _ggseg3d_. Overall, the _ggseg_-packages facilitate parcellation-based visualizations in R, improve and ease the dissemination of the results, and increase the efficiency of the workflows. 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    citation_package: natbib
    fig_caption: true
    includes:
      in_header: preamble.tex
  rticles::peerj_article: preprint
---

```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, collapse = TRUE, results = 'hold', cache=TRUE, fig.pos = 'H') 
options(tinytex.verbose = TRUE, tinytex.clean = FALSE)

library(kableExtra)
library(tidyverse)
library(plotly) 
library(ggplot2)
library(knitr)
library(ggseg)
library(ggseg3d)

@@

dt <- as_tibble(data(package='ggseg')$results)
dt3 <- as_tibble(data(package='ggseg3d')$results)
dtE <- as_tibble(data(package='ggsegExtra')$results)

dd <-  bind_rows(dt, dt3) %>% 
  bind_rows(dtE) %>% 
  filter(!grepl("pal", Item)) %>% 
  mutate(Type = ifelse(grepl("3d", Item), "Mesh", "Polygon"),
         Item = gsub("_3d", "", Item)) %>% 
  select(Package, Title, Type, Item) %>% 
  spread(Type, Package, fill = "None")  %>% 
  mutate(Citation = case_when(
    grepl("dkt", Item) ~ "\\citet{dkt}",
    grepl("yeo", Item) ~ "\\citet{yeo2011}",
    grepl("aseg", Item) ~ "\\citet{aseg}",
    grepl("desterieux", Item) ~ "\\citet{desterieux}",
    grepl("^chen", Item) ~ "\\citet{chen}",
    grepl("glasser", Item) ~ "\\citet{glasser}",
    grepl("icbm", Item) ~ "\\citet{icbm}",
    grepl("jhu", Item) ~ "\\citet{jhu}",
    grepl("schaefer", Item) ~ "\\citet{schaefer}",
    grepl("tracula", Item) ~ "\\citet{tracula}",
    grepl("^ho", Item) ~ "\\citet{ho}",
    TRUE ~ ""
  ))
rm(dt, dt3, dtE)
options(warnPartialMatchAttr = FALSE,
        warnPartialMatchDollar = FALSE,
        warnPartialMatchArgs = FALSE)
```

@@

Visualization is increasingly important for accurate guidance and interpretation of neuroimaging results, as current research is able to generate a high amount of data and outcomes.
For Magnetic Resonance Imaging (MRI), neuroimaging software provides whole-brain information by using many small units of space (>100.000).
Nonetheless, this data is often grouped and summarized into a limited number of regions using predefined brain parcellation atlases.
Brain parcellations segment the brain into a finite set of meaningful neurobiological components, which reflect one or more brain features either based on structural or connectivity properties (@eickhoff_2018).
The use of brain atlases is widespread as these facilitate interpretation and minimize the amount of data, hence reducing problems with multiple comparisons. 
This enables replicability and data sharing in otherwise computationally expensive analyses, which are often performed in specialized software environments such as R (@R).

MRI data provides good spatial resolution and thus an optimal representation has to respect spatial relationships across regions.
Results from brain atlas analyses are most meaningfully visualized when projected onto a representation of the brain, thus it is desirable that any visual representation takes this relation into account.
The projection of data onto brain representations provides clear points of reference - especially when the reader is unfamiliar with the atlas - eases readability, guides interpretation, and conveys the spatial patterns of the data.
Adopting the grammar of graphics implemented in _ggplot2_ (@ggplot), one can plot neuroimaging data directly in R with several tools such as ggBrain (@ggBrain) and ggneuro (@ggneuro; see neuroconductor -@neuroconductor for curated neuroimaging packages for R).
Yet these tools display whole-brain image files and are not well-suited for representing brain atlas data.

In this tutorial, we introduce two packages for visualizing brain atlas data in R.
The _ggseg_ and _ggseg3d_ – pluss the complimentary _ggsegExtra_ – packages include pre-compiled data sets for different brain atlases that allow for 2D and 3D visualization.
The two-dimensional functionality in _ggseg_ is based on polygons and _ggplot2_-based grammar of graphics (@ggplot), while the 3D functionality in _ggseg3d_ is based on tri-surface mesh plots and _plotly_ (@plotly).

Both packages present compiled data sets, tailored functions that allow brain data integration and plotting, and other minor features such as custom colour palettes.
The data featured in the packages are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DKT; @dkt), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; @aseg), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (@fischl_99, @dale_99, @Fischl2000), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (@amlien_elaboration_2019, @WALHOVD20051261, @Pizzagalli).
The _ggsegExtra_ package contains a collection of precompiled atlases (currently `r dd %>% filter(grepl("ggsegExtra", Mesh) | grepl("ggsegExtra", Polygon)) %>% nrow()` additional atlases) and it is frequently updated.
A summary of all available atlases compatible with the _ggseg_-packages as of `r format(Sys.Date(), "%B %Y")` can be found in Table \@ref(tab:atlasTab).

```{r 'atlasTab', echo=F, warning=FALSE, message=FALSE}
dd %>%
  kable(format="latex", escape = FALSE, booktabs = TRUE,
        caption = "Table of currently available atlases in the ggseg, ggseg3d, and the ggsegExtra R-packages. Polygon and mesh refer to 2D and 3D brain atlas representations, respectively") %>%
  kable_styling(latex_options = c("scale_down"))
```

# Tutorial
This tutorial will introduce the _ggseg_, _ggseg3d_, and _ggsegExtra_ packages and familiarize the reader with the main functions and the general use of the packages.
The tutorial will focus on the two main functions: `ggseg()` for plotting 2D polygons and `ggseg3d()` for plotting 3D brains based on tri-surface mesh plots.

@@

_ggseg_ is the main function for plotting 2D data.
By default, the function automatically plots the DKT atlas (see Figure \@ref(fig:init)).
The `ggseg()`-function is a wrapper for `geom_polygon()` from _ggplot2_, and it can be built upon and combined like any _ggplot_-object.
The image plot consists of a simple brain representation containing no extra information.
Hence, _ggseg_ plots can be easily complemented with any of the available _ggplot2_ features and options.
We recommend users to get familiarized with _ggplot2_ (@ggplot).
The package is currently only available through github, but we expect to submit the _ggseg_-package to The Comprehensive R Archive Network (@cran) in 2020.

```{r 'init', fig.width=3, message=FALSE, fig.cap='By default `ggseg()` will plot the dkt atlas in grey shaded polygons.'}
# remotes::install_github("LCBC-UiO/ggseg")
library(ggseg)
library(dplyr)
library(tidyr)

@@

In addition to the standard options for _ggplot2_ polygon geoms, the function also has several options for plotting the main brain representations.
These options are atlas-specific.
For cortical atlases, such as the `dkt`, one can stack the hemispheres, display only the medial or lateral side, choose either one or both hemispheres, or any combination of hemisphere and view (see Figure \@ref(fig:collection) for examples).
For subcortical atlases, such as the `aseg` atlas, the options are more limited but one can often choose between axial, sagittal, and coronal views.

```{r 'collection', fig.width=5, fig.cap='*ggseg* plots can be used with any *ggplot2* feature such as standard scales and themes. For cortical atlases, one can supply special *ggseg* options to determine, for example, hemisphere, view, or position. **A:** `dkt` atlas, stacked with dark theme ; **B:** `dkt` with medial view only; **C:** `dkt` atlas with only left medial display; **D:** `dkt` atlas, stacked, with classic theme; **E:** `dkt` atlas with left hemisphere only; **F:** complete `aseg` atlas'}
# dkt dark theme
p1 <- ggseg(position = "stacked") +
  theme_dark() +
  labs(title=" ")

# dkt classic theme
p2 <- ggseg(position = "stacked") +
  theme_classic() +
  labs(title = " ")

# dkt medial view
med <- ggseg(view = "medial") +
  labs(title = " ")

# dkt left hemisphere
left <- ggseg(hemisphere = "left") +
  labs(title = " ")

# aseg default theme
p3 <- ggseg(atlas=aseg) +
  labs(title = " ")

# dkt left medial alone
combo <- ggseg(view = "medial",
               hemisphere = "left") +
  labs(title=" ")

# Combine plots to Figure 2
cowplot::plot_grid(p1, med, combo,  p2, left, p3,
                   labels = c("A: dkt - dark", "B: dkt - medial",
                              "C: dkt - combo", "D: dkt - classic",
                              "E: dkt - left", "F: aseg"),
                   hjust = -.05)
```

### Using own data with fill and colour
`ggseg()` accepts any argument you can supply to `geom_polygon()`, and therefore is easy to work with for those familiar with _ggplot2_ functionality.
Standard arguments like `fill` that floods the segments with a colour, or `colour` that colours the edges around the segments are typical arguments to provide to the function either as a single setting value or within the _ggplot2_ mapping function `aes`.
To use color palettes corresponding to those used in the original neuroimaging softwares one can use atlas-specific 'brain' palette scales (Figure \@ref(fig:fill)).

```{r 'fill', fig.width=5, fig.height=5, fig.cap="Supplying 'area' to the fill option in `ggseg()`, will use the column 'area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas. The `dkt` atlas default palette corresponds to the FreeSurferColorLut scheme."}
# Figure 3
ggseg(mapping=aes(fill = area), colour="black") +
  scale_fill_brain("dkt") +
  theme(legend.justification=c(1,0),
        legend.position="bottom",
        legend.text = element_text(size = 5)) +
  guides(fill = guide_legend(ncol = 3))
```

Most users will use `ggseg()` to display - using a color scale - some descriptive or inferential statistics, such as mean thickness or brain-cognition relationships across the different brain regions.
Yet, before projecting the statistics onto the segments, one should explore the structure of the atlas data sets.
The atlas data set structure will help users understand what incoming statistical data needs to look like.
Note that each atlas corresponds to a unique data set.
All data sets have a similar structure and contain key information regarding the atlas, the region names, and the coordinates for the segment polygons.

```{r, size='small', eval=FALSE}
# Look at the top 5 rows of the dkt dataset
head(dkt, 5)
```

In any atlas, the column 'label' is particularly useful for combining the data of interest with the _ggseg_-polygons.
The column 'label' contains the label (region) names as in the original neuroimaging software.
For example, the DKT atlas label column matches the region names from Freesurfer statistics table outputs.
Yet, the data in _ggseg_ is in a long format - that is each region has a single row - and any data of interest needs to be in this same format.
Often data sets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged to work with ggseg.
See below an example of wide-to-long conversion.

```{r, size='small'}
# Create a mock freesurfer output
freesurfer_stats <- data.frame(
  id = c(10:12),
  lh_superiortemporal = c(3.32, 4.1, 3.5),
  lh_precentral = c(2.3, 2.5, 2.1),
  lh_rostralmiddlefrontal = c(3.3, 3.2, 3.1)
)
freesurfer_stats

# Wrangle wide format data into long format
freesurfer_long <- freesurfer_stats %>%
  gather(label, thickness, -id)
freesurfer_long
```

Data in long format can then be used directly with the `ggseg()`-function, as the 'label' column corresponds in name and content with the 'label' column in the atlas data of `dkt`.
The data **must** include a column that has the same column name and at least _some_ data matching the values in the corresponding column in the atlas data.
In the next example we create some data with 4 rows, with an 'area' and 'p' column, representing the results of a hypothetical analysis.
The `ggseg()`-function will recognise the matching column 'area', and merge the supplied data into the atlas using _dplyr_-joins.
We use the 'p' column as the column flooding the segment with colour.
The appearance of the plot can then be modified similarly to any other _ggplot2_ graph using functions such as scales, labs, themes, etc., as seen in Figure \@ref(fig:datasupp)

```{r 'datasupp', fig.width=5, fig.cap="Supplying data through the '.data' option in `ggseg()` enables use of columns in the supplied data to aesthetical arguments (such as 'fill'). The *ggseg* plot can be used with any other polygon compatible function from *ggplot2* or *ggplot2* extentions, for instance adding title, changing the legend name and the colour scheme with standard *ggplot2* functions."}
# Make some mock data
someData = data.frame(
  area = c("transverse temporal", "insula",
           "pre central","superior parietal"),
  p = sample(seq(0,.5,.001), 4),
  stringsAsFactors = FALSE)

# Figure 4
ggseg(.data=someData, mapping=aes(fill=p))  +
  labs(title="A nice plot title", fill="p-value") +
  scale_fill_gradient(low="firebrick",high="goldenrod")
```

If the results are only in one hemisphere, but you still want to plot both of them, make sure your data.frame includes the column 'hemi' with either 'right' or 'left'.
In this case, data will be merged into the atlas both by 'area' and by 'hemi'.
For more information about adapting data and viewing only one hemisphere or side, the [package vignettes](https://lcbc-uio.github.io/ggseg/articles/ggseg.html#single-hemisphere-results) contains more elaborate information.

### Creating subplots
There is often the need to plot a statistic of interest in different groups (e.g. thickness or brain - cognition relationships in young or older adults).
This may be obtained also with `ggseg()`, using _ggplot2_'s `facet_wrap` or `facet_grid`, with three guiding rules: 
**1)** as before, data needs to be in long format with a column indexing which group the row corresponds to (group data should appear in seperate rows, not in separate columns).
**2)** The data needs to be grouped using _dplyr_'s `group_by()` function _before_ providing the data to the `ggseg()`-function.
The `ggseg()`-function will detect grouped data, and adapt it to `facet`'s requirements. 
**3)** Apply `facet_wrap` or `facet_grid` to the plot having used the above two rules. 
An example of this can be seen in Figure \@ref(fig:datasupp3), where a mock data set including summary statistics for two groups ('Young' and 'Old') is used when faceting a _ggseg_-plot.

All the concepts described above also work with the `aseg` atlas for subcortical structures, except for 'hemisphere' and 'view' arguments that are superfluous in subcortical atlases (Figure \@ref(fig:atlases)).

```{r 'datasupp3', fig.width=5, fig.cap='To visualize data corresponding to different groups, one needs to use `group_by` or similar **before** plotting with `ggseg()` for facetting to work.'}
# Make some mock data
someData = data.frame(
  area = rep(c("transverse temporal", "insula",
               "pre central","superior parietal"),2),
  p = sample(seq(0,.5,.001), 8),
  AgeG = c(rep("Young",4), rep("Old",4)),
  stringsAsFactors = FALSE) %>%
  group_by(AgeG)

# Figure 5
ggseg(.data=someData, colour="white", position = "stacked",
      mapping=aes(fill=p)) +
  facet_wrap(~AgeG, ncol=2) +
  theme(legend.position = "bottom")
```

```{r 'atlases', fig.width=5, fig.cap='The `aseg` atlas showing subcortical structures has some distinct differences from the `dkt`. For instance, there is no option to show only a single hemisphere. Furthermore, rather than showing lateral and medial surfaces, it shows an axial and sagittal slice.'}
# Figure 6
ggseg(atlas="aseg", mapping=aes(fill=area)) + 
  theme(legend.justification=c(1,0),
        legend.position="bottom",
        legend.text = element_text(size = 5)) +
  guides(fill = guide_legend(ncol = 3))
```

>>>>

---
title: Visualisation of Brain Statistics with R-packages _ggseg_ and _ggseg3d_
preprint: true
author: 
  - name: Athanasia M. Mowinckel
    affiliation: 1
    corresponding: true
    email: a.m.mowinckel@psykologi.uio.no
  - name: Didac Vidal-Piñeiro
    affiliation: 1
affiliation:
  - code: 1
    address: Center for Lifespan Changes in Brain and Cognition, University of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway
abstract: >
  There is an increased emphasis on visualizing neuroimaging results in more intuitive ways. Common statistical tools for dissemination, such as bar charts, lack the spatial dimension that is inherent in neuroimaging data. Here we present two packages for the statistical software R, _ggseg_ and _ggseg3d_, that integrate this spatial component. The _ggseg_ and _ggseg3d_ packages visualize pre-defined brain segmentations as both 2D polygons and 3D meshes, respectively. Both packages are integrated with other well-established R-packages, allowing great flexibility. In this tutorial, we present the main data and functions in the _ggseg_ and _ggseg3d_ packages for brain atlas visualization. The main highlighted functions are able to display brain segmentation plots in R. Further, the accompanying _ggsegExtra_-package includes a wider collection of atlases, and is intended for community-based efforts to develop more compatible atlases for _ggseg_ and _ggseg3d_. Overall, the _ggseg_-packages facilitate parcellation-based visualizations in R, improve and ease the dissemination of the results, and increase the efficiency of the workflows. 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    citation_package: natbib
    fig_caption: true
    includes:
      in_header: preamble.tex
  rticles::peerj_article: preprint
keep_tex: true
---

```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, 
                      collapse = TRUE, results = 'hold', 
                      cache=FALSE, fig.pos = 'H') 
options(tinytex.verbose = TRUE, tinytex.clean = FALSE,
        tibble.width = 50)

library(dplyr)
library(tidyr)
library(plotly) 
library(ggplot2)
library(ggseg)
library(ggseg3d)
library(ggsegExtra)
library(kableExtra)
library(knitr)

@@

get_data_list <- function(package){
  as_tibble(data(package=package)$results)
}

pkglist <- c("ggseg", "ggseg3d",  
             basename(ggsegExtra::ggseg_atlas_repos()$repo))
dt <- lapply(pkglist, get_data_list) %>% 
  bind_rows() %>% 
    filter(!grepl("pal", Item)) %>% 
  mutate(Type = ifelse(grepl("3d", Item), "Mesh", "Polygon"),
         Item = gsub("_3d", "", Item),
         Item = gsub("_", "", Item)) %>% 
    select(Package, Title, Type, Item) %>% 
  spread(Type, Package, fill = "None")  %>% 
  mutate(Citation = case_when(
    grepl("dk", Item) ~ "\\citet{dk}",
    grepl("yeo", Item) ~ "\\citet{yeo2011}",
    grepl("aseg|hcpa", Item) ~ "\\citet{aseg}",
    grepl("desterieux", Item) ~ "\\citet{desterieux}",
    grepl("^chen", Item) ~ "\\citet{chen}",
    grepl("glasser", Item) ~ "\\citet{glasser}",
    grepl("icbm", Item) ~ "\\citet{icbm}",
    grepl("jhu", Item) ~ "\\citet{jhu}",
    grepl("schaefer", Item) ~ "\\citet{schaefer}",
    grepl("tracula", Item) ~ "\\citet{tracula}",
    grepl("^ho", Item) ~ "\\citet{ho}",
    TRUE ~ ""
  ))

options(warnPartialMatchAttr = FALSE,
        warnPartialMatchDollar = FALSE,
        warnPartialMatchArgs = FALSE)
```

@@

Visualization is increasingly important for accurate guidance and interpretation of neuroimaging results, as current research is able to generate a high amount of data and outcomes.
For Magnetic Resonance Imaging (MRI), neuroimaging software provides whole-brain information by using many small units of space (>100.000).
Nonetheless, this data is often grouped and summarized into a limited number of regions using predefined brain parcellation atlases.
Brain parcellations segment the brain into a finite set of meaningful neurobiological components, which reflect one or more brain features either based on structural or connectivity properties (@eickhoff_2018).
The use of brain atlases in neuroscience is widespread as these facilitate interpretation and minimize the amount of data (in comparison to using whole-brain data), hence reducing problems with multiple comparisons.
This enables replicability and data sharing in otherwise computationally expensive analyses, which are often performed in specialized software environments such as R (@R).

MRI data provides good spatial resolution and thus an optimal representation has to respect spatial relationships across regions.
Results from brain atlas analyses are most meaningfully visualized when projected onto a representation of the brain, thus it is desirable that any visual representation takes this relation into account.
The projection of data onto brain representations provides clear points of reference - especially when the reader is unfamiliar with the atlas - eases readability, guides interpretation, and conveys the spatial patterns of the data.
Adopting the grammar of graphics implemented in _ggplot2_ (@ggplot), one can plot neuroimaging data directly in R with several tools such as _ggBrain_ (@ggBrain) and _ggneuro_ (@ggneuro; see neuroconductor -@neuroconductor for curated neuroimaging packages for R).
Yet these tools display whole-brain image files and are not well-suited for representing brain atlas data.

In this tutorial, we introduce two packages for visualizing brain atlas data in R.
The _ggseg_ and _ggseg3d_ – plus the complimentary _ggsegExtra_ – packages, which include pre-compiled data-sets for different brain atlases that allow for 2D and 3D visualization.
The two-dimensional functionality in _ggseg_ is based on polygons and _ggplot2_-based grammar of graphics (@ggplot), while the 3D functionality in _ggseg3d_ is based on tri-surface mesh plots and _plotly_ (@plotly).

The _ggseg_ and _ggseg3d_ packages present compiled data-sets, tailored functions that allow brain data integration and plotting, and other minor features such as custom colour palettes.
The data featured in the packages are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DK; @dk), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; @aseg), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (@fischl_99, @dale_99, @Fischl2000), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (@amlien_elaboration_2019, @WALHOVD20051261, @Pizzagalli).
The _ggsegExtra_ package contains a list of online repositories of pre-compiled atlases (currently `r sum(!grepl("None", dt$Polygon))-1` additional polygon atlases and `r sum(!grepl("None", dt$Mesh))-1` mesh atlases) and it is frequently updated.
A summary of all available atlases compatible with the _ggseg_-packages as of `r format(Sys.Date(), "%B %Y")` can be found in Table \@ref(tab:table1).

```{r 'table1', echo=F, warning=FALSE, message=FALSE}
dt %>%
  kable(format="latex", escape = FALSE, booktabs = TRUE,
        caption = "Table of currently available atlases in ggseg and ggseg3d, and listed in the ggsegExtra R-package. Polygon and Mesh refer to 2D and 3D brain atlas representations, respectively. Cell contents in the Mesh and Polygon columns indicate the source package containing the atlas data-set.") %>%
  kable_styling(latex_options = c("scale_down"))
```

# Tutorial
This tutorial will introduce the _ggseg_, _ggseg3d_, and _ggsegExtra_ packages and familiarize the reader with the main functions and the general use of the packages.
The tutorial will focus on the two main functions: `ggseg()` for plotting 2D polygons and `ggseg3d()` for plotting 3D brains based on tri-surface mesh plots.
The following tutorial complements the package documentation and vignettes that can be found online for all three packages: [ggseg](https://lcbc-uio.github.io/ggseg), [ggseg3d](https://lcbc-uio.github.io/ggseg3d), [ggsegExtra](https://lcbc-uio.github.io/ggsegExtra).

@@

`ggseg()` from the _ggseg_ package is the main function for plotting 2D data.
By default, the function automatically plots the DK atlas (see Figure \@ref(fig:figure1)).
The `ggseg()`-function is a wrapper for `geom_polygon()` from _ggplot2_, and it can be built upon and combined like any _ggplot_-object.
The image plot consists of a simple brain representation containing no extra information.
Hence, _ggseg_ plots can be easily complemented with any of the available _ggplot2_ features and options.
We recommend users to get familiarized with _ggplot2_ (@ggplot).
The package is currently only available through github, but we expect to submit the _ggseg_-package to The Comprehensive R Archive Network (@cran) in 2020.

```{r 'figure1', fig.width=3, message=FALSE, fig.cap='By default `ggseg()` will plot the dk atlas in grey shaded polygons.'}
# remotes::install_github("LCBC-UiO/ggseg")
library(ggseg)

# package combining plots easily
library(patchwork)

@@

In addition to the standard options for _ggplot2_ polygon geoms, the function also has several options for plotting the main brain representations.
These options are atlas-specific.
For cortical atlases, such as the `dk`, one can stack the hemispheres, display only the medial or lateral side, choose either one or both hemispheres, or any combination of hemisphere and view (see Figure \@ref(fig:figure2) for examples).
For subcortical atlases, such as the `aseg` atlas, the options are more limited but one can often choose between axial, sagittal, and coronal views.

(ref:figure2-cap) *ggseg* plots can be used with any *ggplot2* feature such as standard scales and themes. For cortical atlases, one can supply special *ggseg* options to determine, for example, hemisphere, view, or position. **A:** `dk` atlas, stacked with dark theme ; **B:** `dk` with medial view only; **C:** `dk` atlas with only left medial display; **D:** `dk` atlas, stacked, with classic theme; **E:** `dk` atlas with left hemisphere only; **F:** axial view of `aseg` atlas

```{r 'figure2', fig.width=6, fig.cap='(ref:figure2-cap)', out.width = "100%"}
# A: dk dark theme
fig2A <- ggseg(position = "stacked") +
  theme_dark() +
  labs(title="dk", subtitle = "dark theme")

# B: dk medial view
fig2B <- ggseg(view = "medial") +
  labs(title = "dk", subtitle = "medial view")

# C: dk left medial alone
fig2C <- ggseg(view = "medial",
            hemisphere = "left") +
  labs(title="dk", subtitle = "medial left")

# D: dk classic theme
fig2D <- ggseg(position = "stacked") +
  theme_classic() +
  labs(title = "dk", subtitle = "classic theme")

# E: dk left hemisphere
fig2E <- ggseg(hemisphere = "left") +
  labs(title = "dk", subtitle = "left hemisphere")

# F: aseg default theme
fig2F <- ggseg(atlas = aseg, view = "axial") +
  labs(title = "aseg", subtitle = "axial view")

# Using patchwork to combine all plots into one
(fig2A | fig2B | fig2C) / (fig2D | fig2E | fig2F) +
  plot_annotation(tag_levels = 'A')
```

### Using own data with fill and colour
`ggseg()` accepts any argument you can supply to `geom_polygon()`, and therefore is easy to work with for those familiar with _ggplot2_ functionality.
Standard arguments like `fill` that floods the segments with a colour, or `colour` that colours the edges around the segments are typical arguments to provide to the function either as a single setting value or within the _ggplot2_ mapping function `aes`.
To use color palettes corresponding to those used in the original neuroimaging softwares one can use atlas-specific 'brain' palette scales (Figure \@ref(fig:figure3)).

(ref:figure3-cap) Supplying 'region' to the fill option in `ggseg()`, will use the column 'region' from the accompanying data-set to create a discrete colour palette over the segments in the atlas. The `dk` atlas default palette corresponds to the FreeSurferColorLut scheme.

```{r 'figure3', fig.width=5, fig.height=5, fig.cap="(ref:figure3-cap)"}
# Figure 3
ggseg(mapping = aes(fill = region), colour="black") +
  scale_fill_brain("dk") +
  theme(legend.justification = c(1,0),
        legend.position = "bottom",
        legend.text = element_text(size = 5)) +
  guides(fill = guide_legend(ncol = 3))
```

Most users will use `ggseg()` to display - using a color scale - some descriptive or inferential statistics, such as mean thickness or brain-cognition relationships across the different brain regions.
Yet, before projecting the statistics onto the segments, one should explore the structure of the atlas data-sets.
The atlas data-set structure will help users understand what incoming statistical data needs to look like.
Note that each atlas corresponds to a unique data-set.
All data-sets have a similar structure and contain key information regarding the atlas, the region names, and the coordinates for the segment polygons.

```{r, size='small', eval=FALSE, codecap = "Inspect atlas data"}
# Look at the top 5 rows of the dk data-set
head(dk, 5)
```

In any atlas, the column 'label' is particularly useful for combining the data of interest with the _ggseg_-polygons.
The column 'label' contains the label (region) names as in the original neuroimaging software.
For example, the `dk` atlas label column matches the region names from Freesurfer statistics table outputs.
Yet, the data in _ggseg_ is in a long format - that is each region has a single row - and any data of interest needs to be in this same format.
Often data-sets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged to work with ggseg.
Summary statistic files from Freesurfer have a file format that can be difficult to work with in R. 
In a set of functions in _ggseg_, we facilitate conversion between FreeSurfer files and ggseg compatible data. 
There is an [extended vignette](https://lcbc-uio.github.io/ggseg/articles/freesurfer_files.html) of the various options.
In the following example, we import summary data that has been compiled using Freesurfer's `aparcstats2table` (for surfaces) or `asegstats2table` (volumes). 
The summary files can combine parcellation metrics for one or more subjects into a single summary file.
This file also contains extra metrics like intracranial volume and total grey matter volume. 
These later measures do not map onto ggseg, and as such, one should filter them out of the data returned from `read_freesurfer_table)`.

```{r, size='small'}
file_path <- "data/aparc.volume.table"

someData <- read_freesurfer_table(file_path, measure = "volume") %>% 
  # remove rows in the data that do not contain the strings "rh" or "lh"
  dplyr::filter(grepl("rh|lh", label))
someData
```

A second option is to read in atlas files directly from the entire subject directory, using regular expression to select the target atlas. 
The default aparc has a unique string ending with `aparc.stats` which we will use to read in all subject stats for that particular atlas.
One can filter specific subjects and/or regions from the R-object data.frame.

```{r, warning=FALSE}
atlas_data <- read_atlas_files(
  # Path to Freesurfers $SUBJECTS_DIR
  subjects_dir = "/Applications/freesurfer/subjects/",
  # Regular expression to find atlas files
  atlas = "aparc.stats$"
)
atlas_data
```

This long format data can be used directly with the `ggseg()`-function, as the 'label' column corresponds in name and content with the 'label' column in the atlas data of `dk`.
The data **must** include a column that has the same column name and at least _some_ data matching the values in the corresponding column in the atlas data.
In the following example, the `ggseg()`-function will recognise the matching column 'label', and merge the supplied data into the atlas using _dplyr_-joins.
The appearance of the plot can then be modified similarly to any other _ggplot2_ graph using functions such as scales, labs, themes, etc., as seen in Figure \@ref(fig:figure4)

(ref:figure4-cap) Supplying data through the '.data' option in `ggseg()` enables use of columns in the supplied data to aesthetical arguments (such as 'fill'). The *ggseg* plot can be used with any other polygon compatible function from *ggplot2* or *ggplot2* extentions, for instance adding title, changing the legend name and the colour scheme with standard *ggplot2* functions.

```{r 'figure4', fig.width=5, fig.cap="(ref:figure4-cap)", warning = FALSE}

# Figure 4A
fig4A <- someData %>% 
  # plot only subject bert
  filter(subject == "bert") %>%

  ggseg(mapping=aes(fill=volume))  +
  labs(title="from read_freesurfer_table", 
       fill="volume\n(mm^3)") +
  scale_fill_gradient(low="firebrick",high="goldenrod") +
  guides(fill = guide_colourbar(barheight = 3))

# Figure 4B
fig4B <- atlas_data %>% 
  # plot only subject bert
  filter(subject == "bert") %>%

  ggseg(mapping=aes(fill=ThickAvg))  +
  labs(title="from read_atlas_files", 
       fill="Thickness\n(mean)") +
  scale_fill_gradient(low="forestgreen",high="goldenrod")+
  guides(fill = guide_colourbar(barheight = 3))

(fig4A / fig4B) +
  plot_annotation(tag_levels = "A")
```

If the results are only in one hemisphere, but you still want to plot both of them, make sure your data.frame includes the column 'hemi' with either 'right' or 'left'.
In this case, data will be merged into the atlas both by 'region' and by 'hemi'.
For more information about adapting data and viewing only one hemisphere or side, the [online documentation](https://lcbc-uio.github.io/ggseg/articles/ggseg.html#single-hemisphere-results) contains more elaborate information.

### Creating subplots
There is often the need to plot a statistic of interest in different groups (e.g. thickness or brain - cognition relationships in young or older adults).
This may be obtained also with `ggseg()`, using _ggplot2_'s `facet_wrap` or `facet_grid`, with three guiding rules:

1. As before, data needs to be in long format with a column indexing which group the row corresponds to (group data should appear in separate rows, not in separate columns).  
2. The data needs to be grouped using _dplyr_'s `group_by()` function _before_ providing the data to the `ggseg()`-function. 
The `ggseg()`-function will detect grouped data, and adapt it to `facet`'s requirements.  
3. Apply `facet_wrap` or `facet_grid` to the plot having used the above two rules.
An example of this can be seen in Figure \@ref(fig:figure5) B, where a mock data-set including summary statistics for two groups ('Young' and 'Old') is used when faceting a _ggseg_-plot.

The second rule is needed as the `ggseg()`-function is a ggplot-wrapper rather than an extension, and thus has a particular structure and pipeline to combine the atlas to supplied data.
The result of ungrouped data into `ggseg` is shown in  \@ref(fig:figure5) A.
Note that the data is combined incorrectly, there are three panels rather than two. 
Regions not represented in the data (`NA`-values) have their own panel, while the remaining two panels only include regions with valid data.
All the concepts described above also work with the `aseg` atlas for subcortical structures, except for 'hemisphere' and 'view' arguments that are superfluous in subcortical atlases (Figure \@ref(fig:figure6)).

(ref:figure5-cap) Examples of incorrect and correct faceting with `ggseg`. **A**: ungrouped data will facet incorrectly due to the data not being merged correctly with the atlas. **B**: grouped data will merge in the expected way and result in correctly facetted plots.

```{r 'figure5', fig.width=5,  warning= FALSE, fig.cap='(ref:figure5-cap)', out.width="100%"}
# Make some simpler mock data
someData = tibble(
  region = rep(c("transverse temporal", "insula",
               "precentral","superior parietal"),2),
  p = sample(seq(0,.5,.001), 8),
  AgeG = c(rep("Young",4), rep("Old",4))
)

# Figure 5a
fig5A <- someData %>%
  ggseg(position = "stacked",
      mapping=aes(fill=p), show.legend = FALSE) +
  facet_wrap(~AgeG, ncol=3) +
  labs(title = "Ungrouped data")

# Figure 5b
fig5B <- someData %>%
  group_by(AgeG) %>%
  ggseg(position = "stacked",
      mapping=aes(fill=p), show.legend = FALSE) +
  facet_wrap(~AgeG, ncol=3) +
  labs(title = "Grouped data")
